Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-06-22 paper/paper.tex
Further updates to the manuscript.

The above methods can be classified in three categories based on the data models they use. Most of them deal with graphs arising from \emph{aligned sequences}, assuming a shared global sequence with local variation. Methods based on GCSA can index directed acyclic graphs, as well as cyclic graphs that are sufficiently similar to \emph{de~Bruijn graphs}. GCSA2, the hypertext index, and methods based on de~Bruijn graphs work with \emph{arbitrary graphs}.

GCSA2 combines ideas from the original GCSA and from succinct de~Bruijn graphs. Conceptually it uses a de~Bruijn graph as a path index of a variation graph. The \orderk{k} de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely. The pruned graph is then encoded with a generalization of the FM\nobreakdash-index. With two layers of compression, the index often uses less than 1~bit per \kmer{k} for large values of $k$. Similar space usage cannot be achieved with the other de~Bruijn graph-based approaches, which explicitly store information for each \kmer{k}. GCSA2 also includes novel extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph. These extensions are used for e.g.~finding maximal exact matches in the vg read aligner.

@@

\begin{figure}[t!]
\includegraphics[width=396pt]{gcsa2_pruned_index.pdf}
\caption{Left: An \orderk{3} pruned de~Bruijn graph $G''$ \kequivalent{3} to the de~Bruijn graph in Figure~\ref{figure:graph-dbg}. Right: GCSA for graph $G''$. Leftward arrows show how backward searching proceeds, with the highlighted arrows demonstrating it from pattern $\dnaseq{T}$ to pattern $\dnaseq{AT}$. Rightward arrows tell where the samples for each node are located, with the highlighted ones showing them for node $\dnaseq{CAT}$.}\label{figure:pruned-index}
\end{figure}

@@

If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to that node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$. The labels of the predecessor nodes are encoded in $\BWT[sp_{in}, ep_{in}]$. Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v') \in E'$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$. As multiple edges may have the same sort keys, our sorting algorithm must be stable. We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.

@@

If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space. Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use. By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern. Hence we can safely use $\parent$ queries with such ranges.\footnote{The $\parent$ query is similar to the $\textsf{shorter}$ query in the variable-order de~Bruijn graph \cite{Boucher2014}.}

@@

We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space. The resulting index takes less than 16~gigabytes. Query performance is several times slower than with a fast FM\nobreakdash-index for sequences.

>>>>

The above methods can be classified in three categories based on the data models they use. Most of them deal with graphs arising from \emph{aligned sequences}, assuming a shared global sequence with local variation. GCSA-based methods can index directed acyclic graphs, as well as cyclic graphs that are sufficiently similar to \emph{de~Bruijn graphs}. GCSA2, the hypertext index, and indexes based on de~Bruijn graphs work with \emph{arbitrary graphs}.

GCSA2 combines ideas from the original GCSA and from succinct de~Bruijn graphs. Conceptually it uses a de~Bruijn graph as a path index of a variation graph. The \orderk{k} de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely. The pruned graph is then encoded with a generalization of the FM\nobreakdash-index. The index often uses less than 1~bit per \kmer{k} with large values of $k$. Similar space usage cannot be achieved with the other de~Bruijn graph-based approaches, which explicitly store information for each \kmer{k}. GCSA2 also includes novel extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph. These extensions are used for e.g.~finding maximal exact matches in the vg read aligner.

@@

\begin{figure}[t!]
\includegraphics[width=396pt]{gcsa2_pruned_index.pdf}
\caption{Left: An \orderk{3} pruned de~Bruijn graph $G''$ \kequivalent{3} to the de~Bruijn graph in Figure~\ref{figure:graph-dbg}. Right: GCSA for graph $G''$. Leftward arrows show how backward searching proceeds, with the red arrows demonstrating it from pattern $\dnaseq{T}$ to pattern $\dnaseq{AT}$. Rightward arrows tell where the samples for each node are located, with the blue ones showing them for node $\dnaseq{CAT}$.}\label{figure:pruned-index}
\end{figure}

@@

If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to that node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$. The labels of the predecessor nodes are encoded in $\BWT[sp_{in}, ep_{in}]$. Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v') \in E'$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$. As multiple edges may have the same sort key, our sorting algorithm must be stable. We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.

@@

If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range, we can support the required queries in $\Oh{x \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space. By selecting $x$ according to cache line size, the queries require $\Oh{\log_{x} \abs{V'}}$ random memory accesses, which in practice dominate the computation time. Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use. By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern. Hence we can safely use $\parent$ queries with such ranges.\footnote{The $\textsf{shorter}$ query in the variable-order de~Bruijn graph \cite{Boucher2014} is essentially a $\parent$ query.}

@@

We can build a whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space. The resulting index takes less than 16~gigabytes, which is less than what the existing path indexes capable of indexing arbitrary graphs require. Query performance is several times slower than with a similar implementation of the FM\nobreakdash-index for sequences.

GCSA2 improves upon the existing path indexes in the following ways:
\begin{itemize}

\item GCSA \cite{Siren2014} and HISAT2 \cite{Kim2015-2016} use the same techniques, but they can only index graphs that have finite \orderk{\infty} path graphs. On the other hand, they support queries of arbitrary length.

\item BWBBLE \cite{Huang2013} is reported to be 100x slower than BWA \cite{Li2009} and hence tens of times slower than GCSA2. The current implementation can only handle local variation.

\item vBWT \cite{Maciuca2016} is slower than BWBBLE. It can only handle local variation but supports queries of arbitrary length.

\item Succinct \cite{Bowe2012}, compacted \cite{Cazaux2014}, and compressed \cite{Marcus2014} de~Bruijn graphs must store at least one character for each \kmer{k} and hence require more space than GCSA2.

\item The hypertext index \cite{Thachuk2013} has not been implemented. It supports queries of arbitrary length, but becomes slow with matches crossing more than one edge.

\end{itemize}
